Auditory Scene Similarity Retrieval and Classification with Relevance-based Quantization of Scattering Features

### 2016-12-16 paper/structureScene.tex
done with polishing

An important problem in this field is the characterization of auditory scenes for the purposes of similarity estimation. One application of this is acoustic scene similarity retrieval (ASSR), where the scenes judged ``most similar'' to a given query scene are extracted from some larger dataset. This requires the auditory scenes to be represented in some way that captures their distinctive qualities. Often, this has been achieved by bag-of-frames approaches \cite{aucouturier2007bag}, which attempt to describe the auditory scene recording using summary statistics. Unfortunately, since this only captures the average structure of the scene, the approach often fails when presented scenes which are dynamic and contain rare, distinct sound events that discriminate the scene from others. Indeed, experiments in cognitive psychology suggests that this matches more closely the way that humans and other animals perceive acoustic scenes. \gl{put some Refs}

@@

Motivations of the proposed approach and a brief review of the state of the art in the field of acoustic scene modeling is given in Section \ref{sec:soa}. We describe the scattering transform in Section \ref{sec:scattering}, discuss processing of the resulting features in Section \ref{sec:design} and propose a cluster-based scene description in Section \ref{sec:object}. In Section \ref{sec:experiments}, several experiments are proposed for the unsupervised setting of acoustic scene similarity retrieval (ASSR) and for the supervised setting of acoustic scene classification (ASC). Results are subsequently reported in Section \ref{sec:results}.

@@

%The same division is found in the field of auditory scene analysis, wherein the interest may either be 1) to isolate and identify precise events, \ie acoustic event detection (AED), or 2) to predict global attributes assigned to large time spans of audio, \ie acoustic scene classification (ASC).
For these tasks, a popular approach is the bag-of-frames (BOF) approach for acoustic scene modeling introduced by Aucouturier et al. \cite{aucouturier2007bag}, where a scene is modeled by high-level summary statistics computed from local features.
Interestingly, the proposed implementation using Gaussian mixture models (GMMs) of mel-frequency cepstral coefficients (MFCCs) has been demonstrated to perform comparably to a direct averaging of the features for a variety of tasks \cite{lagrange:hal-01082501}.
This contrasts with the typical morphology of acoustic scenes, which is a ``skeleton of events on a bed of textures'' \cite{nelken2013} where a few discrete sound events superimposed upon a stationary acoustic background.
Therefore, it is arguably more beneficial to focus on characterizing these distinct events through describing the large-scale temporal evolution of auditory scenes rather than on the joint probability distribution of short-term features.

@@

Aside from \cite{aucouturier2007bag} and \cite{lagrange:hal-01082501} that consider the task of ASSR, much of the work dedicated to acoustic scene modeling is applied to the ASC task.
An overview of the efforts for the latter task is provided by the IEEE Audio and Acoustic Signal Processing (AASP) Detection and Classification of Acoustic Scenes and Events (DCASE) challenge \cite{barchiesi2015acoustic}, held in 2013 to compare different approaches to scene classification.

@@

\section{Relevance-based quantization of features}
\label{sec:object}

@@

As discussed in Section \ref{sec:soa}, results in sound perception suggest the appropriateness of source-driven representations of auditory scenes for predicting high-level properties. While this can be addressed in the supervised case using late integration of discriminative classifiers (see Section \ref{sec:eili}), this is not possible in the unsupervised case. As the detection of events is still an open problem \cite{7100934}, we consider in this paper a generic quantization scheme in order to identify and represent time intervals of the scene that are coherent, thus likely to be dominated by a given source of interest.

Given a set of $d$-dimensional feature vectors $X_u = \{x_1^u, \ldots, x_L^u\}$, extracted from the scene $s_u$, where $u=\lbrace 1,2,\ldots,U\rbrace$, we would like to partition $X_u$ into a set $C_u = \{c^u_1, \ldots, c^u_M\}$ of $M$ clusters. The partitioning is done by minimizing squared error between the empirical mean, or centroid, of each cluster and the vectors belonging to it. Letting $\mu_m^u$ denote the centroid of the cluster $c_m^u$, we attempt to minimize the following objective function:
\begin{equation}
J(C_u)=\sum\limits_{m} \sum_{x^u_l\in c^u_m} \Vert x_l^u - \mu_m^u \Vert^2\mbox{.}
\end{equation}
This is known as $k$-means clustering.
Each scene $s_u$ is then described by a set of clusters $C_u$. One should note that this quantization approach differs from unsupervised learning schemes such as the ones studied in \cite{bisot2016acoustic}, where the scene features are projected in a dictionary learned from the entire dataset.

%which clusters not over a single scene, but over the whole dataset, and re-projects each scene in the resulting enhanced feature space. \ja{This sentence is also a big unclear.}

Here, with the aim of better balancing the influence of salient sound events and texture-like sounds on the final decision, the similarity between two scenes is computed based on the similarity of their centroids.

@@

@@

The public part of the dataset can used for optimizing the ASC system and the private part used for computing the resulting accuracy using a five-fold cross validation scheme. For this paper, we shall perform our experiments on the private part of the dataset. The folds used are the same as the ones used during the challenge.

The DCASE 2016 dataset has a similar organization, but here we conduct our experiments on the public part of the dataset. \ja{Any more information on this dataset?}

@@

The evaluation is performed on the private part of the DCASE 2013 dataset. The metric used is the precision at rank $k$ ($p@k$), which is computed by taking a query item and counting the number of items of the same class within the $k$ closest neighbors, and then averaging over all query items. The $p@k$ is computed for $k=\lbrace 1,\ldots,9\rbrace$, since each class only has $10$ items. Note that a $p@1$ is equivalent to the classification accuracy obtained by the classifier which chooses the label of the closest neighbor for a given item. For each experimental setting, we only report the results obtained with the sets of parameters, \emph{i.e.} the number of clusters $M$ for the \emph{ob} approaches and the scaling parameter $q$ of the RBF kernels (see Eq. \ref{eq:kc}) leading to the best $p@9$. The \emph{ob} approaches are compared to commonly used early integration approach \emph{early}. \ja{And BOF?}

The clustering for the \emph{ob} methods is done using $k$-means$++$ \cite{arthur2007k}, an augmented version of $k$-means with a more robust initialization procedure. Three numbers of clusters are tested: $8$, $16$ and $32$. The implementation of the presented methods as well as the one of the experimental protocol is available.\footnote{\url{https://github.com/mathieulagrange/paperStructureScene16}} \ja{What is ``the one''?}

@@

The results on acoustic scene classification illustrate the superior performance of the log-scattering transform coupled with late integration, confirming the ideas motivating the construction of our cluster-based auditory scene representation. Using this representation for auditory scene similarity retrieval, we obtain results that improve significantly over traditional BOF and summary approaches, both with log-scattering features and MFCCs.

@@

We report experiments on acoustic scene classification (ASC), following the original methodology of the DCASE 2013 challenge, and extend our approach to its 2016 edition.

@@

\subsubsection*{Participation to the DCASE 2016 challenge}
In order to evaluate the generality of our findings, we have taken part in the acoustic scene classification task of the 2016 edition of the DCASE challenge.
This edition relies on the TUT 2016 acoustic scenes dataset, which contains $15$ classes of acoustic scenes, and is larger than the DCASE 2013 private dataset by one order of magnitude \cite{Mesaros2016_EUSIPCO}.
Training support vector machines with a Gaussian RBF kernel is computationally demanding for a dataset of such size, especially under a late integration scheme, wherein the number of training instances exceeds $10^5$, we therefore use a linear kernel instead.
Our experiments on the smaller DCASE 2013 dataset, as reported in Table \ref{table:dcase2013}, show that the linear kernel gives comparable, albeit slightly worse, classification results than the Gaussian RBF kernel.

@@

@@

This section presents evaluation results for the ASSR settings. The $p@k$ for different settings are shown in Figure \ref{fig:ASS_1}, illustrating the effect of the different similarity metrics.

@@

The same observations are made for the MFCCs for $k\leq5$. However for a rank $k$ greater than $5$, all the \emph{ob} approaches perform equivalently. This may be due to the fact that, at some point, the MFCCs fail to dissociate the discriminative events of a scene from the non-informative ones, thus making the cluster selection/weighting step useless. \ja{What is the ``at some point'' referring to? In other words, why do we expect this to happen for larger $k$ but not smaller?}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{bw/unsupervised_test2}
\caption{Acoustic scene similarity retrieval (ASSR) in the DCASE 2013 private dataset: precisions at rank $k$ ($p@k$) obtained for MFCCs and scattering with logarithmic compression, as a function of the rank $k$.}
\label{fig:ASS_1}
\end{center}
\end{figure}

@@

This paper presents a new approach for modeling acoustic scenes which utilizes scattering transforms at small scales and a cluster-based representation at large scales. Compared to traditional models based on BOF and summary statistics, this representation allows for the characterization of distinct sound events superimposed on a stationary texture, a concept which has strong grounding in the cognitive psychology literature. We demonstrate the usefulness of scattering-based representations for auditory scenes and the value of late integration using experiments in auditory scene classification. In order to extend the ideas of late integration to the unsupervised case, we develop the cluster-based model and validate it using experiments on acoustic scene similarity retrieval. For this task, we show significant improvements over the traditional BOF and summary statistics models based on both standard MFCCs and scattering features.
These outcomes shall be studied further in future work by considering larger databases and more diverse applications, such as pleasantness prediction \cite{lafaySoundscapePilot} as well as other emerging tasks in computational bioacoustics.

>>>>

An important problem in this field is the characterization of auditory scenes for the purposes of similarity estimation. One application of this is acoustic scene similarity retrieval (ASSR), where the scenes judged ``most similar'' to a given query scene are extracted from some larger dataset. This requires the auditory scenes to be represented in some way that captures their distinctive qualities. Often, this has been achieved by bag-of-frames approaches \cite{aucouturier2007bag}, which attempt to describe the auditory scene recording using summary statistics. Unfortunately, since this only captures the average structure of the scene, the approach often fails when presented scenes which are dynamic and contain rare, distinct sound events that discriminate the scene from others. Indeed, experiments in cognitive psychology suggests that this matches more closely the way that humans and other animals perceive acoustic scenes. \gl{put some Refs that are already cited if possible}

@@

Motivations of the proposed approach and a brief review of the state of the art in the field of acoustic scene modeling is given in Section \ref{sec:soa}. We describe the scattering transform in Section \ref{sec:scattering}, discuss processing of the resulting features in Section \ref{sec:design} and propose a cluster-based scene description in Section \ref{sec:object}. In Section \ref{sec:experiments}, several experiments are described for the acoustic scene classification task and the acoustic scene similarity retrieval task. Results are subsequently reported in Section \ref{sec:results}.

@@

%The same division is found in the field of auditory scene analysis, wherein the interest may either be 1) to isolate and identify precise events, \ie acoustic event detection (AED), or 2) to predict global attributes assigned to large time spans of audio, \ie acoustic scene classification (ASC).
In a closely related field, namely the analysis of the urban sound environment, a popular approach is the bag-of-frames (BOF) introduced in the field by Aucouturier et al. \cite{aucouturier2007bag}, where a scene is modeled by high-level summary statistics computed from local features.
Though, recently, the proposed implementation using Gaussian mixture models (GMMs) of mel-frequency cepstral coefficients (MFCCs) has been demonstrated to perform comparably to a direct averaging of the features for a variety of tasks \cite{lagrange:hal-01082501}.
This contrasts with the typical morphology of acoustic scenes, which we believe is a ``skeleton of events on a bed of textures'' \cite{nelken2013} where a few discrete sound events are superimposed upon a stationary acoustic background.
As those distinct and scarce events are most of the time highly relevant for the task at hand, it is arguably more beneficial to focus on characterizing these distinct events through describing the large-scale temporal evolution of auditory scenes rather than on the joint probability distribution of short-term features.

@@

Aside from \cite{aucouturier2007bag} and \cite{lagrange:hal-01082501} that consider the task of acoustic scene similarity retrieval, much of the work dedicated to acoustic scene modeling is applied to the classification task.
An overview of the efforts for the latter task is provided by the IEEE Audio and Acoustic Signal Processing (AASP) Detection and Classification of Acoustic Scenes and Events (DCASE) challenge \cite{barchiesi2015acoustic}, held in 2013 to compare different approaches to scene classification.

@@

\section{Acoustic scene similarity retrieval}
\label{sec:object}

@@

As discussed in Section \ref{sec:soa}, results in sound perception suggest the appropriateness of source-driven representations of auditory scenes for predicting high-level properties. While this can be addressed in the supervised case using late integration of discriminative classifiers (see Section \ref{sec:eili}), this is not direclty feasible in the unsupervised case. As the detection of events is still an open problem \cite{7100934}, we consider in this paper a generic quantization scheme in order to identify and represent time intervals of the scene that are coherent, thus likely to be dominated by a given source of interest.

Given a set of $d$-dimensional feature vectors $X_u = \{x_1^u, \ldots, x_L^u\}$, extracted from the scene $s_u$, where $u=\lbrace 1,2,\ldots,U\rbrace$, we would like to partition $X_u$ into a set $C_u = \{c^u_1, \ldots, c^u_M\}$ of $M$ clusters. The partitioning is done by minimizing squared error between the empirical mean, or centroid, of each cluster and the vectors belonging to it. Letting $\mu_m^u$ denote the centroid of the cluster $c_m^u$, we attempt to minimize the following objective function:
\begin{equation}
J(C_u)=\sum\limits_{m} \sum_{x^u_l\in c^u_m} \Vert x_l^u - \mu_m^u \Vert^2\mbox{.}
\end{equation}
This is known as $k$-means clustering.
Each scene $s_u$ is then described by a set of clusters $C_u$. One should note that this quantization approach differs from unsupervised learning schemes such as the ones studied in \cite{bisot2016acoustic}, where the scene features are projected in a dictionary learned from the entire dataset. Here, with the aim of better balancing the influence of salient sound events and texture-like sounds on the final decision, the similarity between two scenes is computed based on the similarity of their centroids.

@@

The implementations of the presented methods and the experimental protocol are available online\footnote{\url{https://github.com/mathieulagrange/paperStructureScene16}}. % \ja{What is ``the one''?}

@@

The public part of the dataset can used for optimizing the acoustic scene classification system and the private part used for computing the resulting accuracy using a five-fold cross validation scheme. For this paper, we shall perform our experiments on the private part of the dataset. The folds used are the same as the ones used during the challenge.

The DCASE 2016 dataset has a similar organization, with $15$ classes of acoustic scenes, and is larger than the DCASE 2013 private dataset by one order of magnitude \cite{Mesaros2016_EUSIPCO}.

%, but here we conduct our experiments on the public part of the dataset. \ja{Any more information on this dataset?}

@@

The evaluation is performed on the private part of the DCASE 2013 dataset. The metric used is the precision at rank $k$ ($p@k$), which is computed by taking a query item and counting the number of items of the same class within the $k$ closest neighbors, and then averaging over all query items. The $p@k$ is computed for $k=\lbrace 1,\ldots,9\rbrace$, since each class only has $10$ items. Note that a $p@1$ is equivalent to the classification accuracy obtained by the classifier which chooses the label of the closest neighbor for a given item. For each experimental setting, we only report the results obtained with the sets of parameters, \emph{i.e.} the number of clusters $M$ for the \emph{ob} approaches and the scaling parameter $q$ of the RBF kernels (see Eq. \ref{eq:kc}) leading to the best $p@9$. The \emph{ob} approaches are compared to commonly used early integration approach \emph{early}. %\ja{And BOF?}

The clustering for the \emph{ob} methods is done using $k$-means$++$ \cite{arthur2007k}, an augmented version of $k$-means with a more robust initialization procedure. Three numbers of clusters are tested: $8$, $16$ and $32$.

@@

The results on acoustic scene classification illustrate the superior performance of the log-scattering transform coupled with late integration, confirming the ideas motivating the construction of our cluster-based auditory scene representation. Using this representation for auditory scene similarity retrieval, we obtain results that improve significantly over traditional BOF and summary approaches, both with log-scattering and MFCCs features.

@@

We report experiments on acoustic scene classification, following the original methodology of the DCASE 2013 challenge, and extend our approach to its 2016 edition.

@@

\subsubsection*{Participation to the DCASE 2016 challenge}

In order to evaluate the generality of our findings, we have taken part in the acoustic scene classification task of the 2016 edition of the DCASE challenge.
This edition relies on the TUT 2016 acoustic scenes dataset.

Training support vector machines with a Gaussian RBF kernel is computationally demanding for a dataset of such size, especially under a late integration scheme, wherein the number of training instances exceeds $10^5$, we therefore use a linear kernel instead.
Our experiments on the smaller DCASE 2013 dataset, as reported in Table \ref{table:dcase2013}, show that the linear kernel gives comparable, albeit slightly worse, classification results than the Gaussian RBF kernel.

@@

\ml{this section takes one column and is I believe a good candidate for being shrunk or even merged with the previous subsection. By statement like, similar findings ... or those results are confirmed using the 2016 dataset ...}

@@

This section presents evaluation results for the acoustic scene similarity retrieval task. The $p@k$ for different settings are shown in Figure \ref{fig:ASS_1}, illustrating the effect of the different similarity metrics. The metrics is shown for values of $k$ up to 9 which corresponds to the number of elements minus one within each class of scene of the dataset. This allows us to also evaluate the recall capabilities of the system under evaluation.

@@

The same observations are made for the MFCCs for $k\leq5$. However for a rank $k$ greater than $5$, all the \emph{ob} approaches perform equivalently. As such, one can conclude that the use of MFCCs allows us to retrieve closely related scenes, but their lack of discriminative capabilities making the cluster selection/weighting step useless for not so closely related scenes of the same class

%This may be due to the fact that, at some point, the MFCCs fail to dissociate the discriminative events of a scene from the non-informative ones, thus making the cluster selection/weighting step useless. \ja{What is the ``at some point'' referring to? In other words, why do we expect this to happen for larger $k$ but not smaller?}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=.9\columnwidth]{bw/unsupervised_test2}
\caption{Acoustic scene similarity retrieval in the DCASE 2013 private dataset: precisions at rank $k$ ($p@k$) obtained for MFCCs and scattering with logarithmic compression, as a function of the rank $k$.}
\label{fig:ASS_1}
\end{center}
\end{figure}

@@

This paper presents a new approach for modeling acoustic scenes which utilizes scattering transforms at small scales and a cluster-based representation at large scales. Compared to traditional models based on BOF and summary statistics, this representation allows for the characterization of distinct sound events superimposed on a stationary texture, a concept which has strong grounding in the cognitive psychology literature. We demonstrate the usefulness of scattering-based representations for auditory scenes and the value of late integration using experiments in auditory scene classification. In order to extend the selectivity potential of late integration using SVMs to the unsupervised case, we develop the cluster-based model and validate it using experiments on acoustic scene similarity retrieval. For this task, we show significant improvements over the traditional BOF and summary statistics models based on both standard MFCCs and scattering features.
These outcomes shall be studied further in future work by considering larger databases and more diverse applications, such as pleasantness prediction \cite{lafaySoundscapePilot} as well as other emerging tasks in computational bioacoustics.
